Vragenlijst thesis evaluatie

### 2015-06-20 Thesisfollowup/ThesisFollowUpSophieMarien.tex
worm propagatie afgeschreven + related work propagatie

\section{Vergelijk wat je gedaan hebt met het opgestelde werkplan. Hoever sta je? Moet het werkplan aangepast worden?}
Ik moet een tandje bijsteken en al verder beginnen schrijven. Ik was vergeten in mijn werkplan bij te steken dat ik een week op Athens ging gaan. 
\section{Schrijf een eerste versie van een hoofstuk over een afgewerkte bijdrage van je masterproef.}
Dit is meegegeven in bijlage. Ik heb ineens heel mijn voorlopige thesistekst in bijlage gestoken. 
\section{Geef een korte inhoud van het tweede artikel dat je gaat schrijven. (Dit is niet van toepassing voor de studenten van de Master Toegepaste Informatica)}
Het tweede artikel gaat gaan over mijn implementatie van FlipIt op een bedrijfsnetwerk. Eerst bespreek ik hoe ik FlipIt aanpas. Daarna stel ik een methode voor via matrices om de propagatie van een virus te beschrijven. De methode die gebruikt wordt is gelijkaardig aan die van Google Page Ranking. Via deze functie kan ik de totale controle berekenen van een speler over een spel. In het artikel leg ik de methode uit aan de hand van een default spel. 
Ook geef ik een eerste setting van een spel en de bijhorende benefit formules waaruit ik al dan niet een equilibrium kan bepalen.
\section{Beschrijf in 10 lijnen de demo die je bij je verdediging zult geven.}
Ik heb een theoretische thesis dus kan ik moeilijk een demo geven. Ik zal enkele voorbeelden uitwerken om mijn werkingsmethode uit te leggen.
\section{Wat is er afgesproken met je begeleider en/of promotor om op te leveren als major milestone die het praktische werk van je thesis afsluit?}
Dit is nog niet afgesproken.
\section{Evalueer de tijd die je in je thesis gestoken hebt in de voorbije 4 weken: veel te weinig / te weinig / voldoende / ruim voldoende / meer dan verwacht.}
De voorbije weken heb ik minder dan voldoende aan mijn thesis gewerkt. Ik liep vast op een aantal zaken en ik ben een week op Athens cursus geweest. 
\section{Evalueer de vorderingen van je thesis t.o.v. het bereiken van de vooropgestelde doelstelling: ver achter op het schema / achter op het schema / goed op schema / vooruit op het schema / ver vooruit op het schema.}
Ik loop een beetje achter op schema want ik had toch graag nu al een aantal formuleringen bepaald en een model uitgewerkt.
\section{Stuur de ingevulde timesheet mee als bijlage.}

>>>>

Thesis follow up vragen van Maart en April terug beantwoord. 
\section{Vergelijk wat je gedaan hebt met het opgestelde werkplan. Hoever sta je? Moet het werkplan aangepast worden?}
Ik moet een tandje bijsteken en al verder beginnen schrijven. Ik was vergeten in mijn werkplan bij te steken dat ik een week op Athens ging gaan. 
\section{Schrijf een eerste versie van een hoofstuk over een afgewerkte bijdrage van je masterproef.}
Dit is meegegeven in bijlage. Ik heb ineens heel mijn voorlopige thesistekst in bijlage gestoken. 
\section{Geef een korte inhoud van het tweede artikel dat je gaat schrijven. (Dit is niet van toepassing voor de studenten van de Master Toegepaste Informatica)}
Het tweede artikel gaat gaan over mijn implementatie van FlipIt op een bedrijfsnetwerk. Eerst bespreek ik hoe ik FlipIt aanpas. Daarna stel ik een methode voor via matrices om de propagatie van een virus te beschrijven. De methode die gebruikt wordt is gelijkaardig aan die van Google Page Ranking. Via deze functie kan ik de totale controle berekenen van een speler over een spel. In het artikel leg ik de methode uit aan de hand van een default spel. 
Ook geef ik een eerste setting van een spel en de bijhorende benefit formules waaruit ik al dan niet een equilibrium kan bepalen.
\section{Beschrijf in 10 lijnen de demo die je bij je verdediging zult geven.}
Ik heb een theoretische thesis dus kan ik moeilijk een demo geven. Ik zal enkele voorbeelden uitwerken om mijn werkingsmethode uit te leggen.
\section{Wat is er afgesproken met je begeleider en/of promotor om op te leveren als major milestone die het praktische werk van je thesis afsluit?}
Dit is nog niet afgesproken.
\section{Evalueer de tijd die je in je thesis gestoken hebt in de voorbije 4 weken: veel te weinig / te weinig / voldoende / ruim voldoende / meer dan verwacht.}
De voorbije weken heb ik meer dan verwacht aan mijn thesis gewerkt.  
\section{Evalueer de vorderingen van je thesis t.o.v. het bereiken van de vooropgestelde doelstelling: ver achter op het schema / achter op het schema / goed op schema / vooruit op het schema / ver vooruit op het schema.}
Ik loop een beetje achter op schema want ik had toch graag nu al een aantal formuleringen bepaald en een model uitgewerkt.
\section{Stuur de ingevulde timesheet mee als bijlage.}

### 2015-06-20 doc/template/chap-4.tex
versie voor mama
worm propagatie afgeschreven + related work propagatie

\chapter{Virus Propagation }
\label{chapter4: Virus propagation}

\section{Methods of virus propagation}

The spreading of worms has already been extensivly researched [beter schrijven]. Because of all the different types of propagation it is hard to pick out one model that can model all of them. Modelling the spread of malware depends on two great factors: the method used for the propagation and the graph where the malware will spread itself. Viruses and worms are the types of malware that are the most researched. Under we list a couple of examples to model virus and worm propagation. In this text worm and virus will be used intermediate.

\subsection{Models of graph representing the network}

\subsection*{Methods of propagation}

@@

>>>>

\chapter{Worm Propagation }
\label{chapter4: Worm propagation}

The spreading of malware has already been extensively researched. Because of all the different types of propagation it is hard to pick out one model that can model all of them. Modelling the spread of malware depends on two great factors: the method used for the propagation and the graph where the malware will spread itself. Viruses and worms are the types of malware that are the most researched. \\
In this chapter we focus on the propagation models of worms. The propagation of a worm is not depending on the human interaction as it is with viruses. In section [] we present an overview of the most used propagation methods by worms. Next, in section [] different kind of graphs to model a computer network are introduced. Section [] lists all the kind of models to determine the time a worm needs to infect a sufficient number of nodes. These results can be used for the completion of parameter \textit{d}. Finally we introduce an easy method to calculate the delay of the propagation of a worm. 
\subsection{Introduction to this chapter}

Advanced Persistent Threads propagation koppelen met die van een worm.
A virus infects one node on the network and has to wait for human interaction to spread. So the spreading speed depends on the human interaction. 
\subsection*{Methods of propagation}

@@

\todo{toch eigen methode introduceren}

\section{Weetjes}
\begin{description}
\item Crouching Yeti is hardly a sophisticated campaign. For example, the attackers used no zero-day exploits, only exploits that are widely available on the Internet. But that did not prevent the campaign from staying under the radar for several years. \url{http://www.kaspersky.com/internet-security-center/threats/crouching-yeti-energetic-bear-malware-threat}
\end{description}

### 2015-06-20 doc/template/chap-5.tex
worm propagatie afgeschreven + related work propagatie

\subsubsection{Results of the FlipIt game}
The study of the different strategies by means of FlipIt framework allows to derive a number of interesting results:  
\begin{itemize}
\item periodic games dominate the other renewal strategies, meaning that it is always advantageous to play periodically against an opponent with a renewal strategy;
\item periodic games are disadvantageous against players following a Last Move adaptive strategy;
\item if the defender plays with a periodic rate that is fast enough he'll force the attacker to drop out;
\item any amount of feedback about the opponent received during the game, benefits to a player.
\end{itemize}

 In this paper we restrict ourselves to non-adaptive periodic strategies. This choice is motivated by the fact that in a security game a player (defender or attacker) rarely has information about the moves (last move or full history) of his opponent. The results also indicate that the periodic strategy might be the dominant strategy.  \\
\todo{dit stuk tekst hierboven meer motiveren}

>>>>

\subsubsection{Results of the FlipIt game}
The study of the different strategies by means of FlipIt framework allows to derive a number of interesting results:  
\begin{itemize}
\item periodic strategy strongly dominates the other renewal strategies if the opponent has a periodic or non-arithmetic renewal strategy, meaning that it is always advantageous for the opponent to play periodically against a player with a non-adaptive strategy;
\item periodic games are disadvantageous against players following a Last Move adaptive strategy;
\item if a defender facing an LM attacker and with a cost much lower than the attacker, he can play with a periodic rate that is fast enough he'll force the attacker to drop out;
\item a defender facing an LM attacker has to play with a randomized strategy e.g. exponential strategy;
\item any amount of feedback about the opponent received during the game, benefits to a player.
\end{itemize}

% In this paper we restrict ourselves to non-adaptive periodic strategies. This choice is motivated by the fact that in a security game a player (defender or attacker) rarely has information about the moves (last move or full history) of his opponent. The results \cite{FlipIt} also indicate that the periodic strategy might be the dominant strategy.  \\

### 2015-06-20 doc/template/chap-9.tex
worm propagatie afgeschreven + related work propagatie

\subsubsection{Single resource}
The basic FlipIt game consists of a single resource. To represent the security problem of the propagation of an APT in a network, the adapted game defines its single resource as a computer network with multiple
nodes. One of the players, the defender, will also try to defend his network.
% The defender will do this by flipping all the nodes of the network (i.e. the entire resource) in every move he plays. 
The defender will do this by flipping the nodes of the network.
The attacker on the other hand will try to infect all the nodes in the network. 
%For an attacker on the other hand, it is quite impossible to infect all computers in a network at once. Rather, t
The attacker will try to identify the node in the graph that can infect all the nodes in the shortest possible time. Then, after dropping a virus on this first node, it takes a while for the virus to infect the entire network. 
%Since the original FlipIt game works with a single resource that is always flipped entirely, the assumption is made that the attacker is considered to gain immediate full control over the resource when the network has been infected, even it is only one node that has been infected.\\
\subsubsection{Only one action per player}

The action of the defender consist of defending his network. The network consists of multiple nodes. The defender flips all the nodes of the network in every move he plays. The action of the defender can be extended to the action where the defender only flips a subset of the nodes of the network. This extention has already been investigated in the context of placing anti-virus systems in a graph. This problem is known to be NP-hard. [\todo{referentie}].
The defender has only one action. Sending different viruses and worms to a computer network. Every time the attacker flips the resource, he does this by sending a new virus to the system. The attacker does not send the same virus again, because if the computers are patched, the virus cannot get in any more. Sending a virus in this context is equal to flipping a node in the network. The node can be a targeted node or a random node. If the attacker has knowledge about the graph, the attacker can choose to target the node that can infect all the nodes in the shortest time possible.
\subsubsection{Immediate effect of the move}
The framework of FLipIt is used to model targeted attacks. In FlipIt when an attacker attacks, he will get immediately control over the resource. One of the differences is that FlipIt does not consider that moves of the attacker might not be instantaneous. With FlipIt with a propagation delay we model the fact that if the attacker flips, he does not get immediate control over the resource.  This is motivated by the fact that it takes a while before a targeted attack can compromise a computer network. The virus or worm has to spread to infect every computer controlled device in the network. This propagation make take a while. \\
In reality however, after dropping a virus on the first node, it takes a while for the virus to infect
the entire network. So, the assumption that the attacker has full control over the resource as soon as a node has been infected, is not realistic. The attacker has only control of the network once all or a sufficient number of nodes are infected. 

@@

The moves of the defender are immediate. It is assumed that if a defender tries to clean the network that this will happen without a delay. The defender can de-plug the computer from the network, or clean. 
\subsubsection{Stealth character of the move}
Moves in the basic FlipIt game are stealthy or covert and not immediately detected by the other payer. The attacker attacker's moves are stealthy because we want to model a scenario where a computer network is attacker by an APT. The main characteristic of an APT is that the attack is stealthy. Depending on how the attacker has set up his attack, and depending on the network if it is connected with the Internet or not, the moves of the defender are stealthy or not stealthy. For example, if the attacker launches an APT only to harm the system and not to receive feedback, the moves of the defender are stealthy. If the attacker launches an APT to steal sensitive information, the moves of the defender are non stealthy because the attacker can see that he doesn not receive information any more. But the defender can also set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. In thesis we assume that both moves of the attacker and the defender are stealthy. -- kunnen enerzijds niet stealthy zijn als the apt informatie steelt depending how he setup his attacker. depending on network connected. in this a thesis assumed, non stealthy is..
\\
Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some ATP's are designed only to cause harm on the systems that are infected e.g. the Stuxnetworm that was developed to destroy nuclear reactors. The Stuxnetworm was resident on a isolated network, meaning that there was no way to connect to the internet. Another example is by the use of honeypots. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. This is a way how the defender can also be stealthy by providing  false information through the honeypot. 
\subsubsection{Cost associated with the move}
 The defender will be defending every computer controlled device in a network. To make the network clean the defender can patch the systems with the latest security bulletins or reinstall the software or even wipe out the computer. All these different actions can imply other costs. The cost of patching a system will be smaller than the cost of wiping out a computer. \\ 
\todo{cost of the attacker ook toevoegen?}
\subsubsection{Renewal strategies}
The formalization starts from the model of the non-adaptive continuous basic FlipIt game where players use a periodic strategy with a random phase. The basic FlipIt game concluded that periodic strategy is the dominant strategy against all renewal strategies. \todo{zelfde paper conclusie}. Furthermore, this choice is motivated by the assumption that in most organisations, the defence strategy is to periodically defend the network.  This is true for the example of patching. Companies like Microsoft en Google release at a specific moment patches. So companies will apply these patches when they come out. Microsoft security bulletins are released on the Second Tuesday of each month. \cite{MicroPatch}

This corresponds to a periodic defender strategy.  A periodic attacker strategy is assumed as well, to be able to compare the results with the periodic strategy of the FlipIt game in \cite{FlipIt}. %, as this also corresponds to a common real life strategy. %nog eventueel verder te motiveren
Further research can investigate the effect of relaxing this assumption. \\

@@

The worm can use the computers on the network as botnets or steal sensitive data. By patching or installing new software the worm can be removed form the systems. The important assets of the defender to secure in a company network are the computers, servers, data, files, and so fort. In this particular example the resource to defend consists of all the computer controlled devices on the network, e.g. servers and computers. The attacker can drop the worm through a usb stick or by tricking a victim to open or download a malicious program. The last one can be achieved by a trojan horse. 

@@

 A Periodic strategy is a non-adaptive renewal strategy where the time intervals between consecutive moves are a fixed period, denoted by $\delta$. Moreover it has a random phase, that is chosen uniformly and random in the interval $[0,\delta]$ for the first move. The average rate of play of a player is denoted by $\alpha_{i} = \dfrac{1}{\delta_{i}}$. Stated below a list of symbols that will be used throughout the paper. Figure \ref{FlipItDelay} represents the main symbols for clarification.\\
~~\\
\todo{kleuren van de tekening aanpassen zodat ze zichtbaar zijn in een zwartwit paper}
\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/DefFlip.pdf}
\caption{Formalization of a FlipIt game with delay: A representation of a FlipIt game where both players are playing periodically. Every move or flip is indicated by a blue or orange circle, respectively dark gray and light grey.  The defender is represented in blue (dark grey) and plays with a period of $\delta_{D}$. The flip of the attacker is represented by a white circle, but because there is a delay d, the attacker only controls the resource after time d represented by an orange circle (light grey). The attacker plays with a period of $\delta_{A}$. The blue and orange rectangles represent the amount of time the respective player is in control of the resource.}
\label{FlipItDelay}
\end{figure}

>>>>

\subsection{Single resource}
The basic FlipIt game consists of a single resource. To represent the security problem of the propagation of an APT in a network, the adapted game defines its single resource as a computer network with multiple
nodes. One of the players, the defender, will try to defend his network.
% The defender will do this by flipping all the nodes of the network (i.e. the entire resource) in every move he plays. 
The defender will do this by flipping the nodes of the network.
The attacker on the other hand will try to infect all the nodes in the network. 
%For an attacker on the other hand, it is quite impossible to infect all computers in a network at once. Rather, t
%The attacker will try to identify the node in the graph that can infect all the nodes in the shortest possible time. Then, after dropping a virus on this first node, it takes a while for the virus to infect the entire network. \\
The attacker will do this by dropping a virus on a node on the network. The virus will then spread itself and infect other nodes in the network. \\
By defining the resource as a network with multiple nodes it is possible to extend the actions of the defender and the attacker to actions depending on which node is being flipped. These actions will be explained in the following section.
%Since the original FlipIt game works with a single resource that is always flipped entirely, the assumption is made that the attacker is considered to gain immediate full control over the resource when the network has been infected, even it is only one node that has been infected.\\
\subsection{Only one action per player}

The action of the defender consist of defending his network. The network is composed of multiple nodes. The defender can choose to flip one node, a subset of nodes or all nodes of the network. In this paper, the defender flips all the nodes of the network in every move he plays. The action of the defender can be extended to the action where the defender only flips a subset of the nodes or a single node of the network. This extension has already been investigated in the context of placing anti-virus systems in a graph. This problem is known to be NP-hard. [\todo{referentie}]. \\
The defender has only one action. Sending different viruses and worms to a computer network. Every time the attacker flips the resource, he does this by sending a new kind of malware e.g worm to the system. The attacker does not send the same malware again, because if the computers are patched, the malware cannot get in any more. Sending malware in this context is equal to flipping a node in the network. The node can be a targeted node or a random node. If the attacker has knowledge about the graph, the attacker can choose to target the node that can infect all the nodes in the shortest time possible.
\subsection{Immediate effect of the move}
The framework of FLipIt is used to model targeted attacks. In FlipIt when an attacker attacks, he will get immediately control over the resource. One of the differences is that FlipIt does not consider that moves of the attacker might not be instantaneous. With FlipIt with a propagation delay we model the fact that if the attacker flips, he does not get immediate control over the resource.  This is motivated by the fact that it takes a while before a targeted attack can compromise a computer network. The virus or worm has to spread to infect sufficient computer controlled device in the network. This propagation make take a while. \\
%In reality however, after dropping a virus on the first node, it takes a while for the virus to infect
%the entire network. So, the assumption that the attacker has full control over the resource as soon as a node has been infected, is not realistic. The attacker has only control of the network once all or a sufficient number of nodes are infected. 

@@

The moves of the defender are immediate. It is assumed that if a defender tries to clean the network that this will happen without a delay. The defender can de-plug the computer from the network, or just clean a pc. 
\subsection{Stealth character of the move}
Moves in the basic FlipIt game are stealthy or covert and not immediately detected by the other payer. The attacker attacker's moves are stealthy because we want to model a scenario where a computer network is attacker by an APT. The main characteristic of an APT is that the attack is stealthy. Depending on how the attacker has set up his attack, and depending on the network if it is connected with the Internet or not, the moves of the defender are stealthy or not stealthy. For example, if the attacker launches an APT only to harm the system and not to receive feedback, the moves of the defender are stealthy. If the attacker launches an APT to steal sensitive information, the moves of the defender are non stealthy because if the defender takes actions the attacker can see that he does not receive information any more. \\
In paper we assume that both moves of the attacker and the defender are stealthy. Our motivation for a defender with stealthy moves is that it can not be assumed that every attacker can receive feedback from an APT. Some APT's can be resident on a isolated network (e.g. Stuxnetworm), meaning that there was no way to connect to the internet.
%Another example is by the use of honeypots. 
Even when the APT is lauched to steal sensitive information the defender can set-up a honey pot, making the attacker believe that he is stealing sensitive information, but the information in the honeypot is all fake. Honeypots emulate services or creates multiple instances of real operating systems and can pretend to have sensitive information.  Honeypots do not detect all malicious attacks so the attacker can stay unnoticed and can still provide information as feedback. This is a way how the defender can also be stealthy by providing  false information through the honeypot.\\

%  The case where the moves of the defender are non stealthy has already been investigated in a paper of Laszka \citep{MitigationNonTargeted}.

% Some ATP's are designed only to cause harm on the systems that are infected e.g. the Stuxnetworm that was developed to destroy nuclear reactors.

\subsection{Cost associated with the move}
 The defender will be defending every computer controlled device in a network. To make the network clean the defender can patch the systems with the latest security bulletins or reinstall the software or even wipe out the computer. All these different actions can imply other costs. The cost of patching a system will be smaller than the cost of wiping out a computer. \\ 
The cost of the defender depend on the complexity of the vulnerabilities exploited and how difficult it was to program the malware.
\subsection{Strategies}
The formalization starts from the model of the non-adaptive continuous basic FlipIt game where players use a periodic strategy with a random phase. \\

\textbf{No feedback (non-adaptive)}\\
Non-adaptive strategies are taken because of the assumption that both players will receive no feedback during the game. This is motivated by the fact that the identity of the attackers and their attack strategy is rarely known to the defenders. In the case of the attacker, if an attacker wants to have feedback it may be disadvantageous if he wants to stay undetected. \\

\textbf{Periodic strategy}\\
The choice of periodic strategy is motivated by the assumption that in most organisations, the defence strategy is to periodically defend the network.  This is true for the example of patching. Companies like Microsoft en Google release at a specific moment patches. So companies will apply these patches when they come out. Microsoft security bulletins are released on the Second Tuesday of each month. \cite{MicroPatch} \\

In the basic FlipIt game \citep{FlipIt} the authors concluded that periodic strategy is the dominant strategy against all other renewal strategies if the opponent uses a periodic or non-arithmetic renewal strategy.  This means that a good choice for the attacker is also to play a periodic strategy. Furthermore, a periodic attacker strategy is assumed as well, to be able to compare the results with the periodic strategy of the FlipIt game in \cite{FlipIt}.

The paper from Laszka \citep{MitigationNonTargeted}, which was similar to this one, also concluded that periodic is the dominant strategy against all other non-adaptive strategies.
%This corresponds to a periodic defender strategy.   %, as this also corresponds to a common real life strategy. %nog eventueel verder te motiveren
Further research can investigate the effect of relaxing this assumption. \\

%---Periodic is probably the strategy most widely used in practice as most systems require passwords, cryptographic
%keys, etc. to be changed at regular intervals, for example, every thirty days or every three months.
%In [13], it was shown that the periodic strategy strongly dominates all other renewal strategies if the other
%player uses a periodic or non-arithmetic renewal strategy. Thus, the periodic strategy is a good choice for
%an attacker who plays against a non-adaptive (NA) defender.---

@@

The worm can use the computers on the network as botnets or steal sensitive data. By patching or installing new software the worm can be removed form the systems. The important assets of the defender to secure in a company network are the computers, servers, data, files, and so fort. In this particular example the resource to defend consists of all the computer controlled devices on the network, e.g. servers and computers. The attacker can drop the worm through a USB stick or by tricking a victim to open or download a malicious program. The last one can be achieved by a trojan horse. 

@@

 A Periodic strategy is a non-adaptive renewal strategy where the time intervals between consecutive moves are a fixed period, denoted by $\delta$. Moreover it has a random phase, that is chosen uniformly and random in the interval $[0,\delta]$ for the first move. The average rate of play of a player is denoted by $\alpha_{i} = \dfrac{1}{\delta_{i}}$. Stated below a list of symbols that will be used throughout the paper. Figure \ref{FlipItDelay} represents the main symbols for clarification.\\
~~\\

\begin{figure}[hbtp]
\centering
\includegraphics[scale=0.7]{Images/DefFlip.pdf}
\caption{Formalization of a FlipIt game with delay: A representation of a FlipIt game where both players are playing periodically. Every move or flip is indicated by a blue or orange circle, respectively dark gray and light grey.  The defender is represented in blue (dark grey) and plays with a period of $\delta_{D}$. The flip of the attacker is represented by a white circle, but because there is a delay d, the attacker only controls the resource after time d represented by an orange circle (light grey). The attacker plays with a period of $\delta_{A}$. The blue and orange rectangles represent the amount of time the respective player is in control of the resource.}
\label{FlipItDelay}
\end{figure}
